---
title: "PDE Emulation with FFBS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PDE Emulation with FFBS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates how to use `spDBL` to build a statistical emulator for
a spatiotemporal mechanistic model. The target application is a
**Partial Differential Equation (PDE)** solver that maps input parameters to
time-evolving spatial fields. Because running the PDE solver for every new
parameter configuration is expensive, we train a surrogate model on a small
collection of simulator runs and use it to predict the PDE output for unseen
parameter values.

The emulator is built around the **Forward Filter Backward Sampler (FFBS)**
under a Matrix Normal Inverse-Wishart (MNIW) dynamic linear model. The spatial
field is modelled with an autoregressive (AR) structure in time, and a Gaussian
Process (GP) prior captures correlation across the PDE input-parameter space.
Prediction for new parameter values is then carried out via GP-based
interpolation of the smoothed state trajectories.

## Data

The package ships with `dt_emulation`, a list containing training and test
objects from a simulated PDE study:

- `pde_para_train` / `pde_para_test`: matrices of PDE input parameters for the
  training and test sets, respectively.
- `dt_pde_train` / `dt_pde_test`: lists of length `nT` (number of time steps),
  where each element is a matrix of PDE output values over the spatial domain
  for every parameter configuration.

The spatial domain is a 10 × 10 regular grid (`Nx = Ny = 10`, giving
`N_sp = 100` spatial locations). The population size `N_people = 10,000` is
used for rescaling outputs to proportions in the diagnostic plots.

```{r setup}
require(spDBL)
# devtools::load_all() 
require(magrittr)
require(ggplot2)
require(ggpubr)
seed <- 1234
set.seed(seed)

# read  data
data("dt_emulation")
pde_para_train = dt_emulation$pde_para_train
pde_para_test  = dt_emulation$pde_para_test
dt_pde_train   = dt_emulation$dt_pde_train
dt_pde_test    = dt_emulation$dt_pde_test

n_train <- dim(pde_para_train)[1]
n_test <- dim(pde_para_test)[1]
prop_train <- n_train / (n_train + n_test)
nT <- length(dt_pde_train)
nT_ori <- nT

AR_choice <- 2
nsam <- 10
Nx <- 10
Ny <- 10
N_sp <- Nx * Ny
N_people <- 10000
gp_tune <- 0.5
gp_sigma2 <- 1.1
gp_tau2 <- 10^(-4)

bnrow <- n_train #
bnrow_test <- n_test #
bncol <- Ny * 1 #
ind_sp <- data.frame(row = rep(1:Ny, times = Nx), col = rep(1:Nx, each = Ny))


# Initialize para for FFBS
# Get covariance matrix
dist_para <- as.matrix(stats::dist(pde_para_train, method = "euclidean", diag = T, upper = T))
phi_para <- 3 / (gp_tune * max(dist_para))
V_para <- gen_exp_kernel(loc = pde_para_train, phi = phi_para, sigma2 = gp_sigma2, tau2 = gp_tau2) # exponential kernel

ind_sp_EP <- ind_sp[1:bncol, ]
dist_sp <- as.matrix(stats::dist(ind_sp_EP, method = "euclidean", diag = T, upper = T))
phi_sp <- 3 / (gp_tune * max(dist_sp))
# R <- gen_gp_kernel(loc = ind_sp_EP, phi = phi_sp, sigma2 = gp_sigma2, tau2 = gp_tau2)
# R <- sigma_sp[1:bncol, 1:bncol]

# generate season-episode data
V_para_full <- V_para #
Y_full = dt_pde_train #
Y_test_full <- dt_pde_test #

fnrow_train <- n_train #
fnrow_test <- n_test #
fncol <- Nx * Ny
N <- bnrow
S <- bncol
n_block <- fnrow_train / bnrow * fncol / bncol
nT_block <- n_block * nT
# generate index
ind <- generate.grid.rowsnake(fnrow = fnrow_train, fncol = fncol, bnrow = bnrow, bncol = bncol)
ind_test <- generate.grid.rowsnake(fnrow = fnrow_test, fncol = fncol, bnrow = bnrow_test, bncol = bncol)
n_b <- n_block
n_b_test <- dim(ind_test)[1]
Y <- list()
# Y_test <- list()
# V_para <- list()

# set F
if(AR_choice == 1){
  F_ls_train <- gen_F_ls_AR1_EP(nT = nT_ori, n_b = n_b, ind = ind, Y = Y_full)
  F_ls_test <- gen_F_ls_AR1_EP(nT = nT_ori, n_b = n_b, ind = ind_test, Y = Y_test_full)
  # print("using AR1")
} else if(AR_choice == 2){
  F_ls_train <- gen_F_ls_AR2_EP(nT = nT_ori, n_b = n_b, ind = ind, Y = Y_full)
  F_ls_test <- gen_F_ls_AR2_EP(nT = nT_ori, n_b = n_b, ind = ind_test, Y = Y_test_full)
  # print("using AR2")
}


# set Y
for (i in 1:nT) {
  temp_Y <- Y_full[[i]]
  for (j in 1:n_b) {
    Y[[(i-1)*n_b + j]] <- temp_Y[ind[j,2]:ind[j,3], ind[j,4]:ind[j,5]]
  }
}


F_ls <- F_ls_train
V_ls <- V_para
N <- bnrow
S <- bncol
nT_ori <- nT
nT <- nT_block
p <- dim(F_ls[[1]])[2]
G_ls <- diag(p)
W_ls <- diag(p)
n0 <- p + 3
m0 <- matrix(1, nrow = p, ncol = S)
M0 <- diag(p)
D0 <- diag(S)
delta <- 1
# gc()
# print("emulation succeed")
```

### Setup notes

**Episode-partition (EP) blocking.** To keep the MNIW state dimension
manageable, the `N_train × N_sp` data array is partitioned into rectangular
blocks using `generate.grid.rowsnake()`. Each block covers all training
parameter configurations (`bnrow = n_train`) and one column-strip of the
spatial grid (`bncol = Ny`), giving `n_block` blocks per time step.

**AR(2) covariates.** The covariate matrices `F_ls` encode an AR(2) temporal
structure: `gen_F_ls_AR2_EP()` stacks the lag-1 and lag-2 spatial blocks
side-by-side so the state equation at each time step regresses on the two
preceding field values. An identity state-transition matrix `G_ls = I` and
identity noise matrices `W_ls = I` are used; the discount factor `delta = 1`
means no variance inflation is applied.

**Gaussian Process prior on parameters.** `V_para` is the GP covariance matrix
over the `n_train` training parameter configurations, constructed with an
exponential kernel (`gen_exp_kernel()`). This matrix acts as the left-variance
of the MNIW observation noise, encoding the assumption that runs with similar
parameter values produce similar PDE outputs.

## Running FFBS

With the model matrices assembled, we run `FFBS()` to obtain the filtered and
smoothed state distributions, and then draw posterior samples of the latent
state trajectory with `FFBS_sampling()`.

`FFBS()` sweeps forward in time applying the Kalman filter equations under the
MNIW conjugate prior, then sweeps backward to produce smoothed means and
covariances. `FFBS_sampling()` draws `nsam` joint samples of the full state
trajectory from the smoothed posterior.

```{r FFBS}
## FFBS----
time_start <- Sys.time()
# print("Start FFBS")
# print(time_start)
para_ffbs <- FFBS(Y = Y, F_ls = F_ls, G_ls = G_ls,
                  W_ls = W_ls, V_ls = V_ls,
                  m0 = m0, M0 = M0,
                  n0 = n0, D0 = D0,
                  nT = nT, delta = 1.0)
time_end <- Sys.time()
# print("End FFBS")
# print(time_end)
# print("Time FFBS")
# print(time_end - time_start)

# sampling
time_start <- Sys.time()
# print("Start sampling")
# print(time_start)
res_ffbs <- FFBS_sampling(nsam = nsam, para_ffbs = para_ffbs, 
                          F_ls = F_ls, G_ls = G_ls,
                          nT = nT, delta = 1)
time_end <- Sys.time()
# print("End sampling")
# print(time_end)
# print("Time sampling")
# print(time_end - time_start)

# Prediction----
input <- pde_para_train
input_new <- pde_para_test
n_input_new <- dim(input_new)[1]
F_new_ls <- F_ls_test

res_pre_exact_EP <- FFBS_predict_exact(Y = Y, para_ffbs = para_ffbs,
                                       input = input, input_new = input_new,
                                       F_ls = F_ls, F_new_ls = F_new_ls,
                                       nT = nT, phi_para = phi_para, gp_sigma2 = gp_sigma2,
                                       gp_tau2 = gp_tau2, delta = 1.0)
res_pre_MC_EP <- FFBS_predict_MC(nsam = nsam, Y = Y, res_ffbs = res_ffbs,
                                 input = input, input_new = input_new,
                                 F_ls = F_ls, F_new_ls = F_new_ls,
                                 nT = nT, phi_para = phi_para, gp_sigma2 = gp_sigma2,
                                 gp_tau2 = gp_tau2, delta = 1.0)


### transfer back ----
# from episode season to season only
res_pre_exact <- recover_from_EP_exact(dat_EP = res_pre_exact_EP, nT_ori = nT_ori, nT = nT)
res_pre_MC <- recover_from_EP_MC(dat_EP = res_pre_MC_EP, nT_ori = nT_ori, nT = nT, nsam = nsam)
```

### Prediction notes

Two prediction strategies are available:

- **Exact (`FFBS_predict_exact`)**: Conditions on the smoothed state means and
  computes the analytic posterior predictive mean for each test input. This is
  deterministic and fast.
- **Monte Carlo (`FFBS_predict_MC`)**: Propagates the `nsam` sampled state
  trajectories through the GP interpolation, yielding a full sample-based
  predictive distribution that supports uncertainty quantification.

The EP-structured predictions are then reassembled into the original
`nT_ori × N_test × N_sp` format using `recover_from_EP_exact()` and
`recover_from_EP_MC()`.

## Results

### PDE solution over time

The heatmaps below show the true PDE output at nine equally spaced time steps
for the first test parameter configuration. The color scale is shared across all
panels so that temporal dynamics are directly comparable.

```{r plot 1 prepare}
# Plot file ----
# Set ggplot theme
# theme_set(theme_minimal(base_size = 22))
col_epa <- c("#00e400", "#ffff00", "#ff7e00", "#ff0000", "#99004c", "#7e0023")
col_bgr <- c("#d5edfc", "#a5d9f6", "#7eb4e0", "#588dc8", "#579f8b", "#5bb349",
             "#5bb349", "#f3e35a", "#eda742", "#e36726", "#d64729", "#c52429",
             "#a62021", "#871b1c")

## Plot PDE results ----
### Heat Map----

# ind_sp <- data.frame(row = rep(1:Ny, times = Nx), col = rep(1:Nx, each = Ny))
# plot_ls <- list() 
# tstamp <- as.integer(seq(1, nT, length.out = 9))
input_num <- 1
tstamp <- as.integer(seq(1, nT_ori, length.out = 9))
dat <- dt_pde_test
max_y <- max(as.vector(unlist(dat))) # set max limit for all plots
# max_y <- max(as.vector(unlist(res_pre_exact))) # set max limit for all plots

# pde_heat <- plot_panel_heatmap_9(dat = dat, tstamp = tstamp, filename = "plot_panel_pde",
#                                  input_num = input_num, max_y = max_y, Nx = Nx, Ny = Ny, nT = nT_ori, savei = F)
{
  plot_ls <- list()
  ind_sp <- data.frame(row = rep(1:Ny, times = Nx), col = rep(1:Nx, each = Ny))
  ind_plot <- 1

  for (i in tstamp) {
    temp <- dat[[i]][input_num,]
    rownames(temp) <- NULL
    colnames(temp) <- NULL
    dt <- data.frame(row = ind_sp$row, col = ind_sp$col, sol = temp)%>%
      as.data.frame()

    p <- ggplot(dt, aes(x = col, y = row, fill = sol)) +
      geom_raster() +
      scale_fill_gradientn(colours = col_bgr,
                           limits = c(0, max_y),
                           oob = scales::squish) +
      labs(x = "x", y = "y", fill = "Value") 
    # +
    #   theme(text = element_text(size=28),
    #         legend.text = element_text(size = 28),
    #         legend.key.size = unit(1.5, "cm"))

    plot_ls[[ind_plot]] <- p
    ind_plot <- ind_plot + 1
  }
}

pde_heat <- plot_ls
```

```{r plot 1, fig.width=12, fig.height=10, out.width="100%", warning=FALSE, message=FALSE}
labels <- paste0("PDE: t = ", tstamp[1:9] - 1)

ggarrange(
  plotlist = pde_heat[1:9],
  ncol = 3,
  nrow = 3,
  labels = labels,
  font.label = list(size = 14, face = "bold"),
  hjust = -0.1,
  vjust = 1.2,
  align = "hv",
  common.legend = TRUE,
  legend = "right"
)
```

### FFBS emulation vs. PDE solution

The next figure compares the true PDE solution (top row) with the FFBS
emulator's exact posterior predictive mean (bottom row) at three representative
time points. Close visual agreement between the two rows indicates that the
emulator has successfully learned the spatiotemporal dynamics from the training
runs.

```{r plot 2 prepare}
dat <- res_pre_exact # this is exact
# ffbs_heat <- plot_panel_heatmap_9(dat = dat, tstamp = tstamp, filename = "plot_panel_ffbs", savei = F,
#                                   input_num = input_num, max_y = max_y, Nx = Nx, Ny = Ny, nT = nT)
{
  plot_ls <- list()
  ind_sp <- data.frame(row = rep(1:Ny, times = Nx), col = rep(1:Nx, each = Ny))
  ind_plot <- 1

  for (i in tstamp) {
    temp <- dat[[i]][input_num,]
    rownames(temp) <- NULL
    colnames(temp) <- NULL
    dt <- data.frame(row = ind_sp$row, col = ind_sp$col, sol = temp)%>%
      as.data.frame()

    p <- ggplot(dt, aes(x = col, y = row, fill = sol)) +
      geom_raster() +
      scale_fill_gradientn(colours = col_bgr,
                           limits = c(0, max_y),
                           oob = scales::squish) +
      labs(x = "x", y = "y", fill = "Value")
    # +
    #   theme(text = element_text(size=28),
    #         legend.text = element_text(size = 28),
    #         legend.key.size = unit(1.5, "cm"))

    plot_ls[[ind_plot]] <- p
    ind_plot <- ind_plot + 1
  }
}

ffbs_heat <- plot_ls
```

```{r plot 2, fig.width=12, fig.height=7, out.width="100%", warning=FALSE, message=FALSE}
labels <- c(
  paste0("PDE solution: t = ", tstamp[c(4, 6, 8)] - 1),
  paste0("FFBS emulation: t = ", tstamp[c(4, 6, 8)] - 1)
)

ggarrange(
  plotlist = list(
    pde_heat[[4]], pde_heat[[6]], pde_heat[[8]],
    ffbs_heat[[4]], ffbs_heat[[6]], ffbs_heat[[8]]
  ),
  ncol = 3,
  nrow = 2,
  labels = labels,
  font.label = list(size = 16, face = "bold"),
  hjust = -0.1,
  vjust = 1.2,
  align = "hv",
  common.legend = TRUE,
  legend = "right"
)
```

### Uncertainty quantification

The final diagnostic evaluates the MC predictive distribution. Each panel is a
scatter plot where the x-axis shows the true PDE value (rescaled by
`N_people`) and the y-axis shows the FFBS posterior median. Error bars span the
2.5th–97.5th percentile of the `nsam` MC draws. Points falling near the red
diagonal indicate accurate emulation; narrow bars indicate low predictive
uncertainty. The three panels correspond to the same three time points used
above.

```{r plot 3 prepare}
{
  ### Error plot----
  #### specific time, location, all inputs (point)----
  {
    alpha_error <- 0.5
    res_pre <- res_pre_MC
    
    time_num <- 15
    sp_num <- 2
    y_true <- dt_pde_test[[time_num]][,sp_num]
    y_pre <- res_pre[[time_num]][,sp_num,]
    y_pre_stat <- data.frame(y_true = y_true,
                             med = apply(X = y_pre, MARGIN = 1, FUN = median),
                             lower = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.025),
                             upper = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.975))
    error_width <- (max(y_true) - min(y_true)) / 30 # denominator is for aesthetic 
    y_pre_stat %>% ggplot(aes(x = y_true, y = med)) + 
      # geom_linerange(aes(ymin = lower, ymax = upper)) 
      geom_pointrange(aes(ymin = lower, ymax = upper), size =.2)+
      geom_errorbar(aes(ymin = lower, ymax = upper), width = error_width) + 
      geom_abline(col = "red")
  }
  
  #### specific time, input, all locations (field)----
  {
    time_num <- 20
    sp_num <- c(1:N_sp)
    # input_num <- 1
    y_true <- dt_pde_test[[time_num]][input_num,sp_num]
    y_pre <- res_pre[[time_num]][input_num,sp_num,]
    y_pre_stat <- data.frame(y_true = y_true,
                             med = apply(X = y_pre, MARGIN = 1, FUN = median),
                             lower = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.025),
                             upper = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.975))
    y_pre_stat <- y_pre_stat / N_people
    error_width <- (max(y_pre_stat["y_true"]) - min(y_pre_stat["y_true"])) / 30
    plot_error_1 <- y_pre_stat %>% ggplot(aes(x = y_true, y = med)) + 
      # geom_linerange(aes(ymin = lower, ymax = upper)) 
      geom_pointrange(aes(ymin = lower, ymax = upper), size =.2, alpha = alpha_error)+
      geom_errorbar(aes(ymin = lower, ymax = upper), width = error_width, alpha = alpha_error) + 
      geom_abline(col = "red") + 
      labs(x = "PDE solution", y = "FFBS prediction")
  }
  
  # panel
  time_p <- tstamp[c(4, 6, 8)]
  sp_num <- c(1:N_sp)
  plot_error_comp_ls <- list()
  plot_band_ls <- list()
  y_pre_stat_error_comp_ls <- list()
  for (t in 1:length(time_p)) {
    time_num <- time_p[t]
    y_true <- dt_pde_test[[time_num]][input_num,sp_num]
    y_pre <- res_pre[[time_num]][input_num,sp_num,]
    y_pre_stat <- data.frame(y_true = y_true,
                             med = apply(X = y_pre, MARGIN = 1, FUN = median),
                             lower = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.025),
                             upper = apply(X = y_pre, MARGIN = 1, FUN = quantile, prob = 0.975))
    y_pre_stat <- y_pre_stat / N_people
    y_pre_stat_error_comp_ls[[t]] <- y_pre_stat
  }
  
  for (t in 1:length(time_p)) {
    # scatter plot
    error_width <- (max(y_pre_stat_error_comp_ls[[t]]["y_true"]) - min(y_pre_stat_error_comp_ls[[t]]["y_true"])) / 30
    plot_error_1 <- y_pre_stat_error_comp_ls[[t]] %>% ggplot(aes(x = y_true, y = med)) + 
      # geom_linerange(aes(ymin = lower, ymax = upper)) 
      geom_pointrange(aes(ymin = lower, ymax = upper), size =.2, alpha = alpha_error / 3)+
      geom_errorbar(aes(ymin = lower, ymax = upper), width = error_width, alpha = alpha_error / 3) + 
      geom_abline(col = "red") + 
      labs(x = "PDE solution", y = "FFBS emulation")
    plot_error_comp_ls[[t]] <- plot_error_1
    
    # error band plot
    # fsize <- 34
    plot_band_predict <- y_pre_stat_error_comp_ls[[t]] %>%
      ggplot(aes(x = y_true, y = med)) + 
      geom_ribbon(aes(ymin = lower, ymax = upper, x = y_true), fill = "#A6CEE3", alpha = 1) +
      geom_point(aes(y = med), color = "#1F78B4", size = 0.7, alpha = 1) +
      geom_abline(color = "#E31A1C", linewidth = 1) + 
      labs(x = "PDE solution", y = "FFBS emulation") +
      # theme(
      #   axis.title.x = element_text(size = fsize),
      #   axis.title.y = element_text(size = fsize),
      #   axis.text = element_text(size = fsize)  # Adjust axis tick labels as needed
      # ) + 
      theme_minimal()
    plot_band_ls[[t]] <- plot_band_predict
  }
}
```

```{r plot 3, fig.width=13, fig.height=4.5, out.width="100%", warning=FALSE, message=FALSE}
labels <- paste0("t = ", tstamp[c(4, 6, 8)] - 1)

ggarrange(
  plotlist = plot_error_comp_ls[1:3],
  ncol = 3,
  nrow = 1,
  labels = labels,
  font.label = list(size = 16, face = "bold"),
  hjust = -0.1,
  vjust = 1.2,
  align = "hv",
  common.legend = TRUE,
  legend = "right"
)
```

## Summary

This vignette illustrated the full `spDBL` emulation workflow:

1. **Data preparation** — Load training/test PDE runs and set up GP covariance
   matrices for both the parameter space and the spatial domain.
2. **Episode-partition blocking** — Reorganise the data into spatial blocks to
   keep the MNIW state dimension tractable.
3. **AR(2) covariates** — Encode temporal dependence by stacking lag-1 and
   lag-2 spatial blocks as covariates for the dynamic linear model.
4. **FFBS** — Run the forward filter and backward sampler to obtain smoothed
   state distributions over the training parameter configurations.
5. **Prediction** — Interpolate to unseen parameter values using either the
   exact posterior predictive mean or a Monte Carlo sample-based distribution.
6. **Diagnostics** — Validate emulation accuracy through side-by-side heatmaps
   and scatter plots with uncertainty intervals.
